Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses

### 2014-10-17 paper_v1.Rnw
more work on lit. review

Rovai \citeyear{rovai2003search}, combining Tinto's and Bean and Metzner's models with factors specific to online learning and pedagogical styles, proposed a composite persistence model specifically for students in online distance education programs. Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning. And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse. An exploratory study of reasons for attrition suggested the following eight constructs based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}.

@@

>>>>

Rovai \citeyear{rovai2003search}, combining Tinto's and Bean and Metzner's models with factors specific to online learning and pedagogical styles, proposed a composite persistence model specifically for students in online distance education programs. Among other novel factors, Rovai's model acknowledges the critical role of computer literacy in online learning. And yet, while the theoretical models become more developed, they became harder to apply to real-world settings, and hence, the empirical evidence to support them remained sparse. An exploratory study of reasons for attrition suggested the following eight constructs, based on over 1,000 online education students: academic and technical skills, learner motivation, time and support for studies, cost and access to the Internet, and technical problems \cite{muilenburg2005student}. An analysis of student behavior on an online education platform showed that 31\% of variation in achievement could be accounted for by a small set of participation measures \cite{morris2005tracking}.

@@

A survey of over one hundred learners who dropped out of MOOC showed that most learners indicated too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}. A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}. A survival analysis of close to 800 learners who had posted on a MOOC discussion forum suggested that i. the likelihood of dropout was lower for those who actively participated in the first week of the course; ii. those who served as an authority figure in the community on the forum, and iii. those who did not engage in a particular subcommunity on the forum \cite{yang2013turn}. In addition, a number of machine learning approaches to predicting attrition yielded promising results (e.g. \cite{taylor2014likely,halawa2014dropout}), but fell short of offering insight into why a given learner left the course.

### 2014-10-17 paper_v1.tex
more work on lit. review

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college. While piror experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration. Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively. This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2011brief}. Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration. Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.

@@

>>>>

The next generation of psychological models, which were highly influential in the literature, emphasized the critical role of students' ``fit'' in the institution. Tinto's \citeyear{tinto1975dropout} student integration model posited that college students' decision to persist is a function of prior experiences and individual students' characteristics, and experiences during college. While piror experiences and characteristics are fixed, schools can influence the college experience, including the degree of social and academic integration. Tinto operationalized academic and social integration by GPA scores and the frequency of positive interactions with peers and instructors, respectively. This resonates with recent work highlighting the critical role of students' feelings of social belonging in achievement-oriented environments \cite{walton2011brief}.

Tinto's work, which specifically targets traditional college students, prompted universities to be proactive in establishing environments that support student integration. Research on attrition in community college settings reiterates the importance of academic and social integration, but points out that non-persistence could indicate success depending on students' intent---students may leave after accomplishing their goal \cite{bers1991persistence}.

@@

A survey of over one hundred learners who dropped out of MOOC showed that most learners indicated too little time due to work responsibilities, not enough social support inside and outside of the course, and insufficient academic and technical support from the course \cite{gutl2014attrition}. A qualitative analysis of public records, especially forum posts, from 42 MOOCs suggested the following reasons for attrition: lack of time, learner motivation, feelings of isolation, lack of interactivity, insufficient prior knowledge or skills, and hidden costs \cite{khalil2014moocs}. Based on just over one thousand learners who participated in the discussion forum of a particular MOOC, the likelihood of dropout was lower for those who actively participated in the first week of the course, those who served as an authority figure in the community on the forum, and those who did not engage in a particular subcommunity on the forum \cite{rose2014social}.

A number of machine learning approaches to predicting attrition yielded promising results, e.g. \cite{taylor2014likely,halawa2014dropout}, but fell short of offering insight into why a given learner left the course.
